---
source: Rmd
title: "An introduction to binary response variables"
objectives:
  - "Calculate the probabilities of success and failure given binary data."
  - "Interpret the expectation of a binary variable."
  - "Calculate and interpret the odds of success given binary data."
  - "Calculate and interpret the log odds of success given binary data."
keypoints:
questions:
teaching: 10
execises: 10
---
```{r, include=FALSE}
source("../bin/chunk-options.R")
# source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("01-")
library(dplyr)
library(tidyr)
library(ggplot2)
library(latex2exp)
```

In this lesson we will work with binary outcome variables. 
That is, variables which can take on two values.
These could be $0$ vs. $1$, "success" vs. "failure" or "yes" vs. "no".

By analysing binary data, we can estimate the probabilities of success and failure.
For example, we may be interested in the probability that individuals between the 
ages of 55 and 66, who have once been smokers, are still smoking during the NHANES
study. 

The probability of success is estimated by the proportion of individuals who are
still smoking. Similarly, the probability of failure is estimated by the 
proportion of individuals who are no longer smoking.

We calculate these values in RStudio by removing empty rows with `drop_na()`,
subsetting individuals of the appropriate age using `filter()`, counting
the number of individuals in each of the two levels of `SmokeNow` using
`count()` and calculating proportions by dividing the counts by the total number
of non-NA observations using `mutate()`. 


```{r calculate probabilities SmokeNow_Age}
dat %>%
  drop_na(SmokeNow) %>%
  filter(between(Age, 55, 66)) %>%
  count(SmokeNow, name = "n") %>%
  mutate(prop = n/sum(n))
```

We see that the probability of success is estimated as $0.38$ and the probability
of failure is estimated as $0.62$. 

In the linear regression lessons, we modelled the *expectation* of the outcome 
variable, $E(y)$. In the case of binary variables, we will also work with the
expectation of the outcome variable. When $y$ is a binary variable, $E(y)$ is
equal to the proportion of successes. In our example above, $E(y) = \text{Pr}(\text{SmokeNow} = \text{Yes}) = 0.38$. 

> ## Exercise  
> You have been asked to study physical activity (`PhysActive`) 
> in individuals with an FEV1 (`FEV1`) between 3750 and 4250 in the NHANES data.   
> A) Estimate the probabilitites that someone is or is not physically active
> for individuals with an FEV1 between 3750 and 4250.   
> B) What value is $E(\text{PhysActive})$ for individuals 
> with an FEV1 between 3750 and 4250?
>
> > ## Solution
> > A) To obtain the probabilities:
> > 
> > ```{r probabilities PhysActive_FEV1}
> > dat %>%
> >   drop_na(PhysActive) %>%
> >   filter(between(FEV1, 3750, 4250)) %>%
> >   count(PhysActive) %>%
> >   mutate(prop = n/sum(n)) 
> > ```
> > 
> > We therefore estimate the probability of physical activity to be $0.68$
> > and the probability of no physical activity to be $0.32$.  
> > 
> > B) $E(\text{PhysActive}) = Pr(\text{PhysActive} = \text{Yes}) = 0.68$
> {: .solution}
{: .challenge}

Beyond probabilities, binary data is often interpreted through odds. 
The odds are defined as:

$$\frac{E(y)}{1-E(y)}.$$

Since the expectation of $y$ equals the probability of success, the odds can also
be written as:

$$\frac{E(y)}{1-E(y)} = \frac{\text{Pr}(\text{Success})}{1-\text{Pr}(\text{Success})} = \frac{\text{Pr}(\text{Success})}{\text{Pr}(\text{Failure})}.$$

Therefore, an odds greater than $1$ indicates that the probability of success
is greater than the probability of failure. An odds less than $1$ indicates that
the probability of failure is greater than the probability of success. 

As we will see in the next episode, binary outcome variables are modelled through
the *log odds*, i.e.:

$$\text{log}\left(\frac{E(y)}{1-E(y)}\right)$$

As we can see in the plot below, a log odds greater than zero is associated with 
a probability of success greater than 0.5. Likewise, a log odds smaller than 0 
is associated with a probability of success less than 0.5. 

```{r plot log odds vs expectation, echo = FALSE}
tibble(E_y = seq(0,1,length.out = 10000)) %>%
  mutate(odds = E_y / (1 - E_y),
         log_odds = log(E_y / (1 - E_y))) %>%
  ggplot(aes(x = E_y, y = log_odds)) +
  geom_line() +
  xlab(TeX(r'($E(y)$)')) +
  ylab(TeX(r'($log\left(\frac{E(y)}{1-E(y)}\right)$)')) +
  ylim(-6, 6)
```

The interpretation of the probabilities, odds and log odds is summarised in the
table below:

| Measure     | Turning point | Interpretation                                      |
|-------------|---------------|-----------------------------------------------------|
| Probability | 0.5           | Proportion of observations that are successes       |
| Odds        | 1.0           | How many times more likely is success than failure? |
| Log odds    | 0             | If log odds > 0, probability is > 0.5.              |

```{r}
dat %>%
  drop_na(SmokeNow) %>%
  filter(between(Age, 55, 66)) %>%
  count(SmokeNow) %>%
  mutate(prop = n/sum(n)) %>%
  filter(SmokeNow == "Yes") %>%
  summarise(odds = prop/(1-prop),
    log_odds = log(prop/(1-prop)))
```

> ## Exercise  
> You have been asked to study physical activity (`PhysActive`) 
> in individuals with an FEV1 (`FEV1`) between 3750 and 4250 in the NHANES data. 
> Calculate the odds and the log odds of physical activity
> for individuals with an FEV1 between 3750 and 4250. 
> How is the odds interpreted here?
>
> > ## Solution
> > 
> > ```{r odds and log odds PhysActive_FEV1}
> > dat %>%
> >   drop_na(PhysActive) %>%
> >   filter(between(FEV1, 3750, 4250)) %>%
> >   count(PhysActive) %>%
> >   mutate(prop = n/sum(n)) %>%
> >   filter(PhysActive == "Yes") %>%
> >   summarise(odds = prop/(1-prop),
> >     log_odds = log(prop/(1-prop)))
> > ```
> > 
> > Since the odds equal 2.17, we expect individuals with an FEV1 between 3750
> > and 4250 to be 2.17 times more likely to be physically active than not. 
> {: .solution}
{: .challenge}


> ## What does $\text{log}()$ do?
> 
{: .callout}
