---
source: Rmd
title: "Assessing logistic regression fit and assumptions"
objectives:
  - Interpret McFadden's $R^2$ and binned residual plots as assessments of model fit.
  - Assess whether the assumptions of the logistic regression model have been violated. 
keypoints:
  - McFadden's $R^2$ measures relative performance, compared to a model that always predicts the mean. Binned residual plots allow us to check whether the residuals have a pattern and whether particular residuals are larger than expected, both indicating poor model fit. 
  - The logistic regression assumptions are similar to the linear regression assumptions. However, linearity and additivity are checked with respect to the logit of the outcome variable. In addition, homoscedasticity and normality of residuals are not assumptions of binary logistic regression. 
questions:
  - How are McFadden's $R^2$ and binned residual plots interpreted?
  - What are the assumptions of logistic regression?
teaching: 10
execises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
#source("../bin/obtain_data.R")
load("../bin/data.RData")
knitr_fig_path("05-")
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(jtools)
SmokeNow_Age <- dat %>%
  glm(formula = SmokeNow ~ Age,  family = "binomial")
```

In this episode we will check the fit and assumptions of logistic regression models. We will use a pseudo-$R^2$ measure of model fit. Most importantly, we will assess model fit visually using binned residual plots. Finally, we will touch upon the four logistic regression assumptions.

## McFadden's $R^2$ as a measure of model fit
$R^2$ and $R^2_{adj}$ are popular measures of model fit in linear regresssion. 
These metrics can take on values from 0 to 1, with higher values indicating that
more of the outcome variation is accounted for by the dependent variables. 
However, these measures cannot be used in logistic regression. A wide variety
of pseudo-$R^2$ metrics have been developed. We will use McFadden's $R^2$ in
this episode.

McFadden's $R^2$ gives us an idea of the relative performance of our model
compared to a model that predicts the mean. Similarly to the original $R^2$,
McFadden's $R^2$ ranges from 0 to 1, with higher values indicating better 
relative performance. However, by the design of this metric, values close
to 1 are unlikely with real-world data. Therefore, a McFadden's $R^2$ of 0.2 
can already indicate a good relative performance. This metric is returned
by `summ()` from the `jtools` package. 

As with the original $R^2$, this metric should not be used on its own to assess
model fit. We will look at McFadden's $R^2$ alongside binned residual plots
in the next section. 

## Assessing model fit by plotting binned residuals
As with linear regression, residuals for logistic regression can be defined
as the difference between observed values and values predicted by the model. 

Plotting raw residual plots is not very insightful. For example,
let's create residual plots for our `SmokeNow_Age` model. First, we
store the residuals, fitted values and explanatory variable in 
a `tibble` named `residualData`. 
Notice that inside `resid()`, we specify `type = response`. Also note
that `fitted()` returns fitted values on the probability scale. 
Next, we create plotting objects `p1` and `p2`, which will contain
residuals vs. fitted and residuals vs. `Age`, respectively.
We plot these together in one region using `+` from the `patchwork` package.

```{r raw residuals}
residualData <- tibble(resid = resid(SmokeNow_Age, type = "response"),
                       fitted = fitted(SmokeNow_Age),
                       age = SmokeNow_Age$model$Age)

p1 <- ggplot(residualData, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.3) +
  xlab("Pr(SmokeNow = Yes)") +
  ylab("Response residuals")

p2 <- ggplot(residualData, aes(x = age, y = resid)) +
 geom_point(alpha = 0.3) +
  xlab("Age") +
  ylab("Response residuals")

p1 + p2

```

Let's plot binned residuals instead. Binned residuals are averages of the residuals
plotted above, grouped by their associated fitted values or values for `Age`.
Binned residual plots can be made with the `binnedplot` from the `arm` package.
Rather than loading the `arm` package with `library()`, we specify the package 
and function in one go using `arm::binnedplot()`. This prevents clashes between
`arm` and packages which we have loaded earlier in the lesson. Unfortunately,
`binnedplot()` does not work with `patchwork`. To create side-by-side plots,
we will use the command `par(mfrow = c(1,2))` ahead of our plots.

Inside `binnedplot()`, we specify the x and y axes, as well as x and y axis labels.
We specify `main = ""` to suppress the default plot titles. 

```{r binned residuals}
par(mfrow = c(1,2))

arm::binnedplot(x = residualData$fitted,
                y = residualData$resid,
                xlab = "Pr(SmokeNow = Yes)",
                ylab = "Average response residuals",
                main = "") 

arm::binnedplot(x = residualData$age,
                y = residualData$resid,
                xlab = "Age",
                ylab = "Average response residuals",
                main = "")

```

The outer lines on the plot indicate the bounds within which the binned residuals
would be expected to fall, if the model provided a good fit to the data. There
are three things to notice in these plots:

* For relatively low and high probabilities of success, the average binned residuals are more negative than would be expected with a good fit. The same goes for some lower and higher values of `Age`.
* For some probabilities towards the centre of the plot, as well as for some ages towards the centre of the plot, the residuals are more positive than would be expected with a good fit. 
* The residuals appear to have a parabolic pattern. 

Recall that a parabolic pattern can sometimes be resolved by squaring an
explanatory variable. Squaring `Age` indeed reduces the parabolic pattern: 

```{r binned residuals after squaring}
SmokeNow_Age_SQ <- dat %>%
  glm(formula = SmokeNow ~ Age + I(Age^2),  family = "binomial")

residualData <- tibble(resid = resid(SmokeNow_Age_SQ, type = "response"),
                       fitted = fitted(SmokeNow_Age_SQ),
                       age = SmokeNow_Age_SQ$model$Age)

par(mfrow = c(1,2))

arm::binnedplot(x = residualData$fitted,
                y = residualData$resid,
                xlab = "Pr(SmokeNow = Yes)",
                ylab = "Average response residuals",
                main = "") 

arm::binnedplot(x = residualData$age,
                y = residualData$resid,
                xlab = "Age",
                ylab = "Average response residuals",
                main = "")
```

Notice that we are still left with some average binned residuals which suggest poor fit. 
This may be unsurprising, as smoking habits are likely influenced by a lot more than `Age` alone. 

At this point, we can also take a look at McFadden's $R^2$ in the output from `summ()`. This comes
at 0.14, which is in line with the moderate fit suggested by the binned residuals. 

```{r summ SmokeNow_Age_SQ}
summ(SmokeNow_Age_SQ)
```


```{r, eval = FALSE, echo = FALSE}
SmokeNow_Age <- dat %>%
  glm(formula = SmokeNow ~ Age + I(Age^2),  family = "binomial")


binnedplot(x = predict(SmokeNow_Age),
           y = resid(SmokeNow_Age)) 

#residuals vs fitted on probability scale
binnedplot(x = SmokeNow_Age$fitted.values,
           y = resid(SmokeNow_Age, type = "working"),
           xlab = "Pr(SmokeNow = Yes)") 

PhysActive_FEV1 <- dat %>%
 drop_na(PhysActive) %>%
 glm(formula = PhysActive ~ FEV1, family = "binomial")

#residuals vs fitted on probability scale
binnedplot(x = PhysActive_FEV1$fitted.values,
           y = resid(PhysActive_FEV1, type = "response"),
           xlab = "Pr(PhysActive = Yes)") 

PhysActive_FEV1_SQ <- dat %>%
 drop_na(PhysActive) %>%
 glm(formula = PhysActive ~ FEV1 + I(FEV1^2), family = "binomial")

#residuals vs fitted on probability scale
binnedplot(x = PhysActive_FEV1_SQ$fitted.values,
           y = resid(PhysActive_FEV1_SQ, type = "response"),
           xlab = "Pr(PhysActive = Yes)") 


binnedplot(x = predict(PhysActive_FEV1),
           y = resid(PhysActive_FEV1, type = "pearson")) 

binnedplot(x = PhysActive_FEV1$model$FEV1,
           y = resid(PhysActive_FEV1),
           xlab = "FEV1")

ggplot(tibble(resid = resid(PhysActive_FEV1),
              fitted = predict(PhysActive_FEV1)),
              aes(x = fitted, y = resid)) +
  geom_point()

```

>## Exercise
> Create a binned residual plot for our `PhysAvtive_FEV1` model. Then answer 
> the following questions:  
> A) Where along the predicted probabilities does our model tend to overestimate or underestimate the probability of success compared to the original data?  
> B) What pattern do the residuals appear to follow?  
> C) Apply a transformation to resolve the pattern in the residuals. Then, create a new binned residuals plot to show that the pattern has been reduced.  
> D) What is McFadden's $R^2$ for this new model? What does it suggest?
> > ## Solution
> > A) Our model overestimates the probability of physical activity in two of the bins below the 0.4 probability. Our model underestimates the probability of physical activity in two of the bins around the probabilities of 0.5 and 0.65. See the binned residual plot below:
> > ```{r PhysActive FEV1 model}
> > PhysActive_FEV1 <- dat %>%
> >  drop_na(PhysActive) %>%
> >  glm(formula = PhysActive ~ FEV1, family = "binomial")
> > 
> > arm::binnedplot(x = PhysActive_FEV1$fitted.values,
> >            y = resid(PhysActive_FEV1, type = "response"),
> >            xlab = "Pr(PhysActive = Yes)") 
> > ```
> > B) There appears to be a parabolic pattern to the residuals.  
> > C) Adding a squared `FEV1` term resolves most of the parabolic pattern:
> > 
> > ```{r PhysActive FEV1 squared model}
> > PhysActive_FEV1_SQ <- dat %>%
> >  drop_na(PhysActive) %>%
> >  glm(formula = PhysActive ~ FEV1 + I(FEV1^2), family = "binomial")
> > 
> > arm::binnedplot(x = PhysActive_FEV1_SQ$fitted.values,
> >            y = resid(PhysActive_FEV1_SQ, type = "response"),
> >            xlab = "Pr(PhysActive = Yes)") 
> > ```
> > D) Since McFadden's $R^2$ is 0.03, it suggests that FEV1 is not a strong predictor of physical activity. 
> > ```{r summ PhysActive_FEV1_SQ}
> > summ(PhysActive_FEV1_SQ)
> > ```
> > 
> {: .solution}
{: .challenge}

## Assessing the assumptions of the logistic regression model
The assumptions underlying the logistic regression model are similar to those
of the simple linear regression model. The key similarities and differences are:  
* Validity, representativeness and independent errors are assessed in the same way. See [this episode](https://carpentries-incubator.github.io/simple-linear-regression-public-health/06-fitAndAssumptionSLR/index.html) from the simple linear regression lesson for explanations and exercises for these assumptions.
* While logistic regression has the linearity and additivity assumption, it is slightly different. This assumption states that *the logit of* our outcome variable has a linear, additive relationship with the explanatory variables. Violations of the linearity assumption can sometimes be solved through transformation of explanatory variables. Violations of the additivity assumption can sometimes be solved through the inclusion of interactions.
* Homoscedasticity and normally distributed residuals are *not* assumptions underlying the logistic regression model.



